---
title: "Predict compounds mechanism-of-action by morphological profiling"
author: "Allen Goodman and Shantanu Singh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predict compounds mechanism-of-action by morphological profiling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates predicting compound mechanism-of-action using
morphological profiling data. See the vignette `single_cell_analysis` for 
details about this dataset.

```{r libraries, message=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(stringr)
library(cytominrworkshop)
```

# Load data

Per-well profiles computed in `single_cell_analysis` are loaded, as well as  
metadata associated with these profiles (obtained from [BBBC021](https://data.broadinstitute.org/bbbc/BBBC021/))

```{r message=FALSE}
profiles <- 
  readr::read_csv(system.file("extdata", "ljosa_jbiomolscreen_2013_per_well_mean.csv", 
                package = "cytominrworkshop"))

moa <-
  readr::read_csv(system.file("extdata", "BBBC021_v1_moa.csv", 
                              package = "cytominrworkshop")) %>%
  rename(Image_Metadata_Compound = compound,
                Image_Metadata_Concentration = concentration,
                Image_Metadata_MoA = moa
  )

metadata <-
  readr::read_csv(system.file("extdata", "BBBC021_v1_image.csv", 
                              package = "cytominrworkshop")) %>%
  rename(Image_Metadata_Plate = Image_Metadata_Plate_DAPI,
                Image_Metadata_Well = Image_Metadata_Well_DAPI
  ) %>%
  select(matches("^Image_Metadata")) %>%
  inner_join(moa) %>%
  distinct()

profiles %<>%
  inner_join(metadata)

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")

```

How many compounds?

```{r}
profiles %>% 
  filter(Image_Metadata_Compound != "DMSO") %>%
  distinct(Image_Metadata_Compound) %>%
  tally() %>%
  rename(`Number of compounds` = n) %>%
  knitr::kable()
```

How many unique treatments (compound-concentration pairs)?

```{r}
profiles %>% 
  filter(Image_Metadata_Compound != "DMSO") %>%
  distinct(Image_Metadata_Compound, Image_Metadata_Concentration) %>%
  tally() %>%
  rename(`Number of unique treatments` = n) %>%
  knitr::kable()
```

How many replicates per unique treatment?

```{r}
profiles %>% 
  filter(Image_Metadata_Compound != "DMSO") %>%
  count(Image_Metadata_Compound, Image_Metadata_Concentration) %>%
  rename(`Number of replicates` = n) %>%
  knitr::kable()
```

How many DMSO wells per plate?

```{r}
profiles %>% 
  filter(Image_Metadata_Compound == "DMSO") %>%
  count(Image_Metadata_Plate) %>%
  rename(`Number of DMSO wells` = n) %>%
  knitr::kable()
```

# Feature selection

Next, lets filter the set of features based on various measures of quality 

Remove features that have near-zero variance. This dataset doesn't have any 
such features, so nothing is removed.

```{r message=FALSE}
profiles <-
  cytominr::select(
    population = profiles,
    variables = variables,
    sample = profiles,
    operation = "variance_threshold"
  ) %>%
  collect()

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")
```


Remove features that have poor correlation across replicates. To do so, lets 
first compute the correlations.

```{r message=FALSE}
doParallel::registerDoParallel(cores = 2)

feature_replicate_correlations <-
  profiles %>% 
  variable_quality(
    variables = variables,
    strata = c("Image_Metadata_Compound", "Image_Metadata_Concentration"),
    replicates = 3
  )
```

What the does the distribution look like? 

```{r fig.width=4, fig.height=4, message=FALSE}
ggplot(feature_replicate_correlations, aes(median))  + 
  stat_ecdf() + 
  geom_vline(xintercept = 0.5, color = "red") + 
  xlab("median replicate correlation (Pearson)") +
  ylab("F(x)")
```

Here, we select a threshold and remove features that have a replicate 
correlation lower than that threshold

```{r message=FALSE}

profiles %<>% 
  select_(.dots = setdiff(x = colnames(profiles), 
                          y = feature_replicate_correlations %>% 
                            filter(median < 0.5) %>% 
                            magrittr::extract2("variable"))
          )

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")
```


Filter based on correlation between features. The morphological features 
extracted contain several highly correlated groups. We want to to prune the set
of features, retaining only one feature from each of these highly correlated
sets. The function `correlation_threshold` provides an approximate (greedy)
solution to this problem. After excluding the features, no pair of features 
have a correlation greater than `cutoff` indicated below.

```{r message=FALSE}
profiles <-
  cytominr::select(
    population = profiles,
    variables = variables,
    sample = profiles,
    operation = "correlation_threshold",
    cutoff = 0.95) %>%
  collect()

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")
```


# Normalize with reference to DMSO

There may be plate-to-plate variations, which can be compensated for to some
extent by normalizing the features with respect to the DMSO wells per plate. 


```{r message=FALSE}

profiles <-
  cytominr::normalize(
    population = profiles,
    variables = variables,
    strata =  c("Image_Metadata_Plate"),
    sample = profiles %>% filter(Image_Metadata_Compound == "DMSO")
  )

profiles <-
  cytominr::select(
      population = profiles,
      variables = variables,
      operation = "drop_na_columns"
  )

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")

```


# Calculate treatment profiles and predict mechanism-of-action

We have selected features and normalized the data. We can now compute 
treatment provides by averaging across replicates.

```{r message=FALSE}

profiles <-
  cytominr::aggregate(
    population = profiles,
    variables = variables,
    strata = c("Image_Metadata_Compound", 
               "Image_Metadata_Concentration",
               "Image_Metadata_MoA"),
    operation = "mean"
  )
```


Let's visualize this data using t-SNE. 

```{r fig.width=8, fig.height=6, message=FALSE}

profiles %<>% 
  filter(Image_Metadata_Compound != "DMSO") 
  
correlation <- 
  profiles %>%
  select(one_of(variables)) %>% 
  as.matrix() %>% 
  t() %>%
  cor()

mechanism <- as.character(profiles$Image_Metadata_MoA)

set.seed(123)

df <- 
  tibble::as_data_frame(
    tsne::tsne(as.dist(1-correlation))
    ) %>% 
  mutate(mechanism = mechanism)

p <- 
  ggplot(df, aes(V1, V2, color=mechanism)) + 
  geom_point() + 
  ggtitle("t-SNE visualization of compound profiles")

print(p)

```

The data clusters into mechanisms quite nicely. Let's quantify this 
by evaluating how well we can predict mechanism-of-action by simply assigning 
a treatment the mechanism of its nearest neighbor.
NOTE: A common mistake when analyzing this dataset is to not exclude other 
concentrations of the same compound when looking up the nearest neighbor. That is cheating! The `mask` addresses this. 

```{r message=FALSE}

compound <- profiles$Image_Metadata_Compound

mask <- as.integer(outer(compound, compound, FUN="!="))

mask[mask == 0] <- -Inf
  
correlation_masked <- correlation * mask

prediction <- sapply(1:nrow(correlation_masked), 
               function(i) mechanism[order(correlation_masked[i,], 
                                           decreasing = TRUE)[1]])

confusion_matrix <- caret::confusionMatrix(prediction, mechanism)
```

What's the classification accuracy?

```{r}
tibble::frame_data(
  ~metric, ~value,
  "Accuracy", sprintf("%.2f", confusion_matrix$overall["Accuracy"]), 
  "95% CI", sprintf("(%.2f, %.2f)", confusion_matrix$overall[["AccuracyLower"]],
                    confusion_matrix$overall[["AccuracyUpper"]])
  ) %>% 
  knitr::kable(digits = 2)
```

What does the whole confusion matrix look like?

```{r}
confusion_matrix$table %>% 
  knitr::kable()

```




