---
title: "Predict compounds mechanism-of-action by morphological profiling"
author: "Tim Becker and Mohammad Rohban"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Improving predicting compounds mechanism-of-action by morphological profiling using SNF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates predicting compound mechanism-of-action using
morphological profiling data. The data is "cleaned up" using the similarity 
network fusion algorithm (SNF). This vignette bases on the vignette 
`predict_moa`. See the vignette `single_cell_analysis` for
details about this dataset.

```{r libraries, message=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(stringr)
library(cytominergallery)
library(SNFtool)
```

# Load data

Per-well profiles computed in `single_cell_analysis` are loaded, as well as
metadata associated with these profiles (obtained from [BBBC021](https://data.broadinstitute.org/bbbc/BBBC021/)).
This is the same data as used in the vignette `predict_moa`.

```{r message=FALSE}
profiles <-
  readr::read_csv(system.file("extdata", "ljosa_jbiomolscreen_2013_per_well_mean.csv",
                package = "cytominergallery"))

moa <-
  readr::read_csv(system.file("extdata", "BBBC021_v1_moa.csv",
                              package = "cytominergallery")) %>%
  rename(Image_Metadata_Compound = compound,
                Image_Metadata_Concentration = concentration,
                Image_Metadata_MoA = moa
  )

metadata <-
  readr::read_csv(system.file("extdata", "BBBC021_v1_image.csv",
                              package = "cytominergallery")) %>%
  rename(Image_Metadata_Plate = Image_Metadata_Plate_DAPI,
                Image_Metadata_Well = Image_Metadata_Well_DAPI
  ) %>%
  select(matches("^Image_Metadata")) %>%
  inner_join(moa) %>%
  distinct()

profiles %<>%
  inner_join(metadata)

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")

```


# Feature selection

Next, lets filter the set of features based on various measures of quality

Remove features that have near-zero variance. This dataset doesn't have any
such features, so nothing is removed.
```{r message=FALSE}
profiles <-
  cytominer::select(
    population = profiles,
    variables = variables,
    sample = profiles,
    operation = "variance_threshold"
  ) %>%
  collect()

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")
```


Remove features that have poor correlation across replicates. To do so, lets
first compute the correlations.

```{r message=FALSE}

doParallel::registerDoParallel(cores = 4)

feature_replicate_correlations <-
  profiles %>%
  cytominer::variable_importance(
    variables = variables,
    strata = c("Image_Metadata_Compound", "Image_Metadata_Concentration"),
    replicates = 3,
    cores = 2)
```


Similar to the `predict_moa` vignette, we select a threshold and remove features that 
have a replicate correlation lower than that threshold.

```{r message=FALSE}

profiles %<>%
  select_(.dots = setdiff(x = colnames(profiles),
                          y = feature_replicate_correlations %>%
                            filter(median < 0.5) %>%
                            magrittr::extract2("variable"))
          )

variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")
```


Filter based on correlation between features, similar to `predict_moa`. The 
morphological features extracted contain several highly correlated groups. 
We want to to prune the set of features, retaining only one feature from each 
of these highly correlated sets. The function `correlation_threshold` provides 
an approximate (greedy) solution to this problem. After excluding the features, 
no pair of features have a correlation greater than `cutoff` indicated below.

```{r message=FALSE}
profiles <-
  cytominer::select(
    population = profiles,
    variables = variables,
    sample = profiles,
    operation = "correlation_threshold",
    cutoff = 0.95) %>%
  collect()


variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")


```

# Normalize with reference to DMSO

There may be plate-to-plate variations, which can be compensated for to some
extent by normalizing the features with respect to the DMSO wells per plate.

```{r message=FALSE}

profiles <-
  cytominer::normalize(
    population = profiles,
    variables = variables,
    strata =  c("Image_Metadata_Plate"),
    sample = profiles %>% filter(Image_Metadata_Compound == "DMSO")
  )

profiles <-
  cytominer::select(
      population = profiles,
      variables = variables,
      operation = "drop_na_columns"
  )



variables <-
  colnames(profiles) %>%
  str_subset("^Nuclei_|^Cells_|^Cytoplasm_")

```


# Calculate treatment profiles and predict mechanism-of-action

We have selected features and normalized the data. We can now compute
treatment profiles by averaging across replicates.

```{r message=FALSE}

profiles <-
  cytominer::aggregate(
    population = profiles,
    variables = variables,
    strata = c("Image_Metadata_Compound",
               "Image_Metadata_Concentration",
               "Image_Metadata_MoA"),
    operation = "mean"
  )
```


Let's visualize this data using t-SNE.

```{r fig.width=8, fig.height=6, message=FALSE}

profiles %<>%
  filter(Image_Metadata_Compound != "DMSO")

correlation <-
  profiles %>%
  select(one_of(variables)) %>%
  as.matrix() %>%
  t() %>%
  cor()

mechanism <- as.character(profiles$Image_Metadata_MoA)


plot_tsne <- function(correlation, mechanism){
  set.seed(42)
  df <-
    tibble::as_data_frame(
      tsne::tsne(as.dist(1 - correlation))
      ) %>%
    mutate(mechanism = mechanism)

  p <-
    ggplot(df, aes(V1, V2, color = mechanism)) +
    geom_point() +
    ggtitle("t-SNE visualization of compound profiles clean using SNF")

  print(p)
}

plot_tsne(correlation = correlation, mechanism = mechanism)
```

As we saw in predict_moa, the data clusters into mechanisms quite nicely. Let's 
quantify this by evaluating how well we can predict mechanism-of-action by simply 
assigning a treatment the mechanism of its nearest neighbor.
NOTE: A common mistake when analyzing this dataset is to not exclude other
concentrations of the same compound when looking up the nearest neighbor. That is 
cheating! `mask` in the code below addresses this.

```{r message=FALSE}

predict_moa <- function(correlation, compound, mechanism){
  mask <- as.integer(outer(compound, compound, FUN = "!="))
  mask[mask == 0] <- -Inf

  correlation_masked <- correlation * mask

  return(sapply(1:nrow(correlation_masked),
                 function(i) mechanism[order(correlation_masked[i,],
                                             decreasing = TRUE)[1]])
    )
}

compound <- profiles$Image_Metadata_Compound
prediction <- predict_moa(correlation, compound, mechanism)
confusion_matrix <- caret::confusionMatrix(prediction, mechanism)
```

What's the classification accuracy?

```{r}

evaluate_prediction <- function(confusion_matrix ){
  tibble::frame_data(
    ~metric, ~value,
    "Accuracy", sprintf("%.2f", confusion_matrix$overall["Accuracy"]),
    "95% CI", sprintf("(%.2f, %.2f)", confusion_matrix$overall[["AccuracyLower"]],
                      confusion_matrix$overall[["AccuracyUpper"]])
    )
}

evaluate_prediction(confusion_matrix) %>%
  knitr::kable(digits = 2)
```

What does the whole confusion matrix look like?

```{r}
confusion_matrix$table %>%
  knitr::kable()

```

Next we test the similarity network fusion (SNF) to clean the data. 

```{r}
distance <- 1-correlation
affinity <- affinityMatrix(distance,  K = 20, sigma = 0.5)
snf_distance <- SNF(list(affinity, affinity), K = 20, t = 20 )
```

Do we get better results using SNF?
```{r}
prediction <- predict_moa(snf_distance, compound, mechanism)
confusion_matrix <- caret::confusionMatrix(prediction, mechanism) 

evaluate_prediction(confusion_matrix) %>%
  knitr::kable(digits = 2)
```

Visualize the results using tsne. 

```{r}

plot_tsne(correlation = snf_distance, mechanism = mechanism)

```

Now the data is split in different feature categories that are combined using SNF. 

```{r}

feature_label = list("_AreaShape_", "_Intensity_","_Texture_","_Neighbors_")

feature_names_list <- lapply(feature_label, function(x) (
  profiles %>%
    select(matches(x)) %>% 
    colnames())
  )

correlation_matrices <- lapply(feature_names_list, function(x) (
  profiles %>%
    select(one_of(x)) %>%
    as.matrix() %>%
    t() %>%
    cor())
)

affinity_list <- lapply(correlation_matrices, function(x) (
  affinityMatrix(1 - x, K = 20, sigma = 0.5 ))
)

snf_distance <- SNF(affinity_list, K = 20, t = 20 )
```

How is the performance?
```{r}

prediction <- predict_moa(snf_distance, compound, mechanism)

confusion_matrix <- caret::confusionMatrix(prediction, mechanism) 

evaluate_prediction(confusion_matrix) %>%
  knitr::kable(digits = 2)
```

Again tsne plot
```{r}

plot_tsne(correlation = snf_distance, mechanism = mechanism)

```
We split the data in different constituents  

```{r}
#profiles %<>%
#  filter(Image_Metadata_Compound != "DMSO")

constituents = list("Cells", "Nuclei","Cytoplasm")

feature_list <- lapply(constituents, function(x)
  (profiles %>%
    select(matches(x)) %>% 
    colnames())
  )

correlation_matrices <- lapply(feature_list, function(x) (
  profiles %>%
  select(one_of(x)) %>%
  as.matrix() %>%
  t() %>%
  cor())
)

affinity_list <- lapply(correlation_matrices, function(x) (
  affinityMatrix(1 - x, K = 20, sigma = 0.5 )
))

snf_distance <- SNF(affinity_list, K = 20, t = 20 )
```


Do we get better results?
```{r}
prediction <- predict_moa(snf_distance, compound, mechanism)

confusion_matrix <- caret::confusionMatrix(prediction, mechanism) 

evaluate_prediction(confusion_matrix) %>%
  knitr::kable(digits = 2)

plot_tsne(correlation = snf_distance, mechanism = mechanism)

```