---
title: "Feature distribution analysis"
output:
  html_notebook: default
  html_document: default
---

This notebook analyzes feature distributions. We categorize features as being multimodal, skewed but unimodel, or symmetric and unimodal. A statistic (Hartigan's dip test statistic) is used to check for multimodality. Unimodal features with skewness > 2 are categorized as "skewed". All other features are categorized as symmetric and unimodal. This is reported in a table in the notebook and also available as a CSV file. Additionally, we plot the histogram of each feature, saved as a PNG.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_libraries, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(magrittr)
library(purrr)
library(readr)
library(stringr)
library(tidyr)
library(tibble)
library(reshape2)
library(rbenchmark)
```


```{r set_seed}

set.seed(42)

```

## Download files

```{sh eval=FALSE}
mkdir -p ~/Downloads/BBBC022_workspace

cd ~/Downloads/BBBC022_workspace

wget https://s3.amazonaws.com/imaging-platform-collaborator/2016_09_09_cytominer_workshop/backend_BBBC022_20646.tar.gz

wget https://s3.amazonaws.com/imaging-platform-collaborator/2016_09_09_cytominer_workshop/metadata_BBBC022.tar.gz

tar xvzf backend_BBBC022_20646.tar.gz

tar xvzf metadata_BBBC022.tar.gz

```

## Load database backend

```{r load_backend}

workspace_dir <- file.path(Sys.getenv("HOME"), "Downloads", "BBBC022_workspace")

batch_id <- "BBBC022_2013"

plate_id <- "20646"

plate_backend <- 
  file.path(workspace_dir, 
            paste0("backend/", batch_id, "/", plate_id, "/", plate_id,".sqlite"))

db <- src_sqlite(path = plate_backend)

```

## Load metadata

```{r load_metadata}

barcode_platemap <- 
  suppressMessages(read_csv(file.path(workspace_dir, paste0("metadata/", batch_id, "/barcode_platemap.csv"))))

metadata <- 
paste0(
  file.path(workspace_dir, paste0("metadata/", batch_id,"/platemap/")),
  barcode_platemap$Plate_Map_Name %>% unique(),
  ".txt"
  ) %>% 
  map_df(function(x) suppressMessages(read_tsv(x))) %>%
  rename(Plate_Map_Name = plate_map_name) %>%
  inner_join(barcode_platemap, by = c("Plate_Map_Name")) %>%
  mutate(Plate = Assay_Plate_Barcode,
         Well = well_position) %>%
  mutate(broad_sample = ifelse(is.na(broad_sample), "DMSO", broad_sample))

names(metadata) %<>% str_c("Metadata", ., sep = "_")

if (db_has_table(db$con, table = "metadata")) {
  db$con %>% db_drop_table(table = "metadata")
}
metadata <- dplyr::copy_to(db, metadata)

```


## Sample cells from DMSO wells

```{r sample_cells_from_dmso}

frac_cells_per_image <- .8

images_per_well <- 6

# sample images from DMSO wells
sampled_images <- 
  metadata %>%
  filter(Metadata_broad_sample == "DMSO") %>% 
  inner_join(tbl(db, "Image"), by = c("Metadata_Plate" = "Image_Metadata_Plate", 
                         "Metadata_Well" = "Image_Metadata_Well")) %>%
  select(matches("Metadata_|TableNumber|ImageNumber")) %>%
  collect() %>%
  group_by(Metadata_Plate, Metadata_Well) %>%
  sample_n(images_per_well) %>%
  ungroup()

if (db_has_table(db$con, table = "sampled_images")) {
  db$con %>% db_drop_table(table = "sampled_images")
}  

sampled_images <- dplyr::copy_to(db, sampled_images)

# sample cells from the sampled images
sampled_objects <-
  sampled_images %>%
  inner_join(
    tbl(db, "Cells") %>% select(TableNumber, ImageNumber, ObjectNumber),
    by = c("TableNumber", "ImageNumber")) %>%
  collect() %>%
  group_by(TableNumber, ImageNumber) %>%
  sample_frac(frac_cells_per_image) %>%
  ungroup()

if (db_has_table(db$con, table = "sampled_objects")) {
  db$con %>% db_drop_table(table = "sampled_objects")
}  

sampled_objects <- dplyr::copy_to(db, sampled_objects)
  
sampled_objects %<>%
  inner_join(tbl(db, "Cells"), by = c("TableNumber", "ImageNumber", "ObjectNumber")) %>%
  inner_join(tbl(db, "Cytoplasm"), by = c("TableNumber", "ImageNumber", "ObjectNumber")) %>%
  inner_join(tbl(db, "Nuclei"), by = c("TableNumber", "ImageNumber", "ObjectNumber")) %>%
  collect(n = Inf)

```


## Peform Hartigan's dip test for multimodality.

```{r diptest}

diptest_p_values <- 
  sampled_objects %>%
  select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>%
  map(function(x) diptest::dip.test(x)[["p.value"]]) %>%
  as_data_frame() %>%
  gather(feature, diptest_p_value)

diptest_p_values$diptest_p_value_adjusted <-
  p.adjust(diptest_p_values$diptest_p_value, method = "BH")

```


## Compute skewness

```{r skewness}

skewnesses <- 
  sampled_objects %>%
  select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>%
  map(e1071::skewness) %>%
  as_data_frame() %>%
  gather(feature, skewness)

```

## Compute feature ranges

```{r feature_ranges}

feature_ranges <- 
  sampled_objects %>%
  select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>%
  gather(feature, value) %>%
  group_by(feature) %>%
  summarise_at(vars(value), 
               funs(min = min(., na.rm = TRUE),
                    q01 = quantile(., probs = 0.01, na.rm = TRUE),
                    q99 = quantile(., probs = 0.99, na.rm = TRUE),
                    max = max(., na.rm = TRUE)))

```

# Collect, summarize and save statistics

```{r features_statistics}

features_statistics <- 
  inner_join(diptest_p_values, skewnesses, by = c("feature")) %>%
  inner_join(feature_ranges, by = c("feature")) %>%
  mutate(is_multimodal = diptest_p_value_adjusted < 0.05) %>%
  mutate(is_skewed = skewness > 2) %>%
  mutate(is_skewed = ifelse(is_multimodal, NA, is_skewed))
  
features_statistics %>%
  group_by(is_skewed) %>%
  tally() %>% 
  knitr::kable(caption = "Feature skewness summary", digits = 2)

features_statistics %>%
  group_by(is_multimodal) %>%
  tally() %>% 
  knitr::kable(caption = "Feature multimodality summary", digits = 2)

features_statistics %>% 
  mutate(neg_log_p_value_adjusted_dip = -log(diptest_p_value_adjusted, base = 10)) %>%
  filter(diptest_p_value_adjusted < 0.25) %>% 
  select(feature, neg_log_p_value_adjusted_dip) %>%
  arrange(-neg_log_p_value_adjusted_dip) %>% 
  knitr::kable(caption = "Feature multimodality table", digits = 3)

features_statistics %>% 
  select(feature, skewness) %>%
  na.omit() %>%
  filter(abs(skewness) > 2) %>% 
  arrange(-abs(skewness)) %>% 
  knitr::kable(caption = "Feature skewness table", digits = 3)

features_statistics %>% 
  select(feature, min, q01, q99, max) %>%
  knitr::kable(caption = "Feature range table", digits = 3)

features_statistics %>% write_csv("feature_statistics.csv")

```

# Plot densities of features

```{r plot_densities}

if(!dir.exists("feature_densities")) {
  dir.create("feature_densities")  
}

feature_names <- names(sampled_objects) %>% 
  str_subset("Cells_|Cytoplasm_|Nuclei_")

for (feature_name in feature_names) {
  g <- ggplot(sampled_objects, aes_string(feature_name)) + 
    stat_density(adjust = 0.8, alpha = 0.5)
  
  ggsave(plot = g, 
         filename = sprintf("feature_densities/%s.png", feature_name), 
         width = 5, height = 5)
}

```

# reporting the correlations using a distributed version of the correlation function

for n.splits <- 32 and n.cores <- 32 on m4.16xlarge AWS instances here is the benchmark for the paralle cor and standard R cor

sampled_objects %>% select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>% cytominergallery::parallel_cor(., n.splits = n.splits, n.cores = n.cores, cov.fun = "custom_multi_covar")
  replications elapsed relative user.self sys.self user.child sys.child
            3   7.356        1     3.768    1.996     70.176     6.253
                                                                         test
sampled_objects %>% select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>% cor
  replications elapsed relative user.self sys.self user.child sys.child
            3   47.54        1     47.35    0.175          0         0

```{r report_correlations}

splits <- 2
cores <- 4
# how many top correlated feature to each given feature is shown
n.top.cor <- 5          

sampled_objects %>%
  select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>% 
  cytominergallery::parallel_cor(splits = splits, 
                                 cores = cores, 
                                 cov_fun = "two_pass_multi_covar") %>% 
  reshape2::melt(value.name = "cor.") %>%
  mutate(feature.1 = Var1, feature.2 = Var2) %>%
  select(-Var1, -Var2) %>%
  arrange(-abs(cor.)) %>%
  filter(feature.1 != feature.2) %>%
  group_by(feature.1) %>%
  do(head(., n.top.cor)) %>%
  ungroup() %>%
  knitr::kable(legend = "Top feature correlations")

# benchmarking the running time of the parallel_cor
rbenchmark::benchmark(sampled_objects %>%
  select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>%
  cytominergallery::parallel_cor(splits = splits, 
                                 cores = cores, 
                                 cov_fun = "two_pass_multi_covar"), replications = 3) %>% 
  print

# benchmarking the running time of the standard cor function
rbenchmark::benchmark(sampled_objects %>%
  select(matches("^Cells_|^Cytoplasm_|^Nuclei_")) %>%
  cor) %>% 
  print

```
